---
title: 'Data Mining Project: Clustering and Playlist Creation'
author: "Stephen Gianfortoni, Anna Whelan, and Scott Shepard"
date: "August 6, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=F}
library(tidyverse)
library(caret)
library(GGally)
library(leaps)
library(mclust)
library(flexmix)
library(tidyLPA)
library(RColorBrewer)
library(hrbrthemes)
library(forcats)
```

## Playlist Generation

__Load Data__

```{r}
music.data <- read.csv('C:/Users/gianf/Documents/MS Analytics/Data Mining/Project_Data.csv')
music.data$listens <- music.data$listens + 2
music.data$song_hotttnesss <- music.data$song_hotttnesss + 0.0001 # so that songs do not have a zero chance of being picked
```

__Exploratory Analysis__

```{r}
head(music.data[, 14:21], 10)

str(music.data)

summary(music.data)
```

__Cluster Generation__

```{r}
set.seed(1027)
split.index <- createDataPartition(music.data$track_id, p = 0.632, 
                                  list = F, 
                                  times = 1)


music.k.means <- vector("list", length = 15)
train <- music.data[split.index, ]
train$tempo <- (train$tempo-min(train$tempo))/(max(train$tempo)-min(train$tempo))


set.seed(1027)
for (i in 1:15) {
    music.k.means[[i]] <- kmeans(train[, 14:21], centers = i, nstart = 100, iter.max = 15)
}

```

__Scree Plot__


```{r}
scree <- data.frame(clusters = rep(NA, 15), within.ss = rep(NA, 15), vaf = rep(NA, 15))
for (i in 1:15) {
    scree$clusters[i] <- i
    scree$within.ss[i] <- music.k.means[[i]]$tot.withinss
    scree$vaf[i] <-  music.k.means[[i]]$betweenss /music.k.means[[i]]$totss
}

ggplot(data = scree, aes(x = clusters, y = vaf)) + geom_line() + geom_point()
```

Based on the Scree plot above. 5-6 clusters appear to be the ideal amount.

__Cluster Center Interpretation__

```{r}
music.k.means[[6]]$centers
```

```{r}
centers.data <- as.data.frame(music.k.means[[6]]$centers)
centers.data$group <- 1:6
centers.data$set <- 'train'

centers.data <- centers.data %>% 
    gather(key = 'variable',
           value = 'value',
           factor_key = TRUE,
           -group,
           -set)

ggplot(centers.data, aes(x = variable, y = value, group = group)) + geom_col(position = 'dodge') + 
    geom_col(position = 'dodge') + facet_grid(~group) + coord_flip() + theme(legend.position = 'bottom')
```

Based on the information above. The 6 clusters could be interpreted as follows:

1. Highly acoustic, highly instrumental. Low energy and dancibility. Low tempo and valence.
2. High energy, low in speech, acousticness, and instrumentalness.
3. Low energy acoustic songs. Higher proportion of liveness and speechiness than others.
4. Dancible, positive, and high energy songs. Mostly instrumental.
5. High energy, instrumental, acoustic songs
6. High energy and high instrumentalness. Low valence (emotionally negative)



__Cluster Stability on Holdout Data__

```{r}
holdout.k.means <- vector("list", length = 15)
test <- music.data[-split.index, ]
test$tempo <- (test$tempo-min(test$tempo))/(max(test$tempo)-min(test$tempo))

set.seed(1027)
for (i in 1:15) {
    holdout.k.means[[i]] <- kmeans(test[, 14:21], centers = as.matrix(music.k.means[[i]]$centers), nstart = 1)
}


for (i in 1:15) {
    scree$holdout.clusters[i] <- i
    scree$holdout.within.ss[i] <- holdout.k.means[[i]]$tot.withinss
    scree$holdout.vaf[i] <-  holdout.k.means[[i]]$betweenss /holdout.k.means[[i]]$totss
}
```

```{r}
ggplot(data = scree, aes(x = clusters)) + geom_line(aes(y = holdout.vaf), col = 'red') + 
    geom_point(aes(y = holdout.vaf), col = 'red') + geom_line(aes(y = vaf), col = 'black') + 
    geom_point(aes(y = vaf), col = 'black')
```

```{r}
print(paste0(scree$holdout.clusters, ': ', scree$holdout.vaf))
```

__Cluster Comparison__

```{r}
print(holdout.k.means[[5]]$centers)
```

```{r}
centers.test <- as.data.frame(holdout.k.means[[6]]$centers)
centers.test$group <- 1:6
centers.test$set <- 'test'

centers.test <- centers.test %>% 
    gather(key = 'variable',
           value = 'value',
           factor_key = TRUE,
           -group,
           -set)

centers.combined <- rbind(centers.data, centers.test)

ggplot(centers.combined, aes(x = variable, y = value, group = interaction(group,set), fill = set)) + 
    geom_col(position = 'dodge') + geom_col(position = 'dodge') + facet_grid(~group) + coord_flip() + 
    theme(legend.position = 'bottom')
```

```{r}
train$cluster <- as.factor(music.k.means[[6]]$cluster)

test$cluster <- as.factor(holdout.k.means[[6]]$cluster)

table(train$cluster, train$genre_top)

prop.table(table(test$cluster))

prop.table(table(music.data$genre_top))
```

Both the centers and the proportions of the songs in each cluster appear to be very stable.

__Generate Playlists - no genre__

```{r}
recombined <- rbind(train, test)

ggplot(music.data, aes(song_hotttnesss)) + geom_density(fill = 'chartreuse3') + ggtitle('Music Hotness Density')
    
set.seed(1027)

group1 <- sample_n(recombined[recombined$cluster == 1,], 25, weight = song_hotttnesss)

group6 <- sample_n(recombined[recombined$cluster == 6,], 25, weight = song_hotttnesss)

group4 <- sample_n(recombined[recombined$cluster == 4,], 25, weight = song_hotttnesss)

group5 <- sample_n(recombined[recombined$cluster == 5,], 25, weight = song_hotttnesss)
```

## Genre Specific Playlists

Do the different genres have differing density plots for the type of music they are?

```{r}
recombined.quant.feat <- recombined[, 11:21]

recombined.quant.feat <- recombined.quant.feat %>% 
    gather(key = 'feature',
           value = 'value',
           factor_key = TRUE,
           -genre_top,
           -interest,
           -song_title)

ggplot(recombined.quant.feat, aes(x = value, y = ..scaled..)) + 
    geom_density(aes(fill = feature)) + facet_grid(feature~genre_top) + 
    theme(legend.position = 'bottom')
```

Based on the plots above, and basic intuition, a hypothesis can be made that different genres should be considered different populations and clustered separately.

```{r}
table(music.data$genre_top)
```

The table above shows that the largest groups in the data set are Hip Hop, Electronic, Folk, and Rock. These genres will be clustered separately.

## Rock Cluster

```{r}
rock <- music.data[music.data$genre_top == 'Rock',]

set.seed(1027)
rock.split.index <- createDataPartition(rock$track_id, p = 0.632, 
                                  list = F, 
                                  times = 1)

rock.train.cluster <- vector("list", length = 15)
rock.train <- rock[rock.split.index, ]
rock.train$tempo <- (rock.train$tempo-min(rock.train$tempo))/(max(rock.train$tempo)-min(rock.train$tempo))


set.seed(1027)
for (i in 1:15) {
    rock.train.cluster[[i]] <- kmeans(rock.train[, 14:21], centers = i, nstart = 100, iter.max = 15)
}


rock.scree <- data.frame(clusters = rep(NA, 15), within.ss = rep(NA, 15), vaf = rep(NA, 15))
for (i in 1:15) {
    rock.scree$clusters[i] <- i
    rock.scree$within.ss[i] <- rock.train.cluster[[i]]$tot.withinss
    rock.scree$vaf[i] <-  rock.train.cluster[[i]]$betweenss /rock.train.cluster[[i]]$totss
}

ggplot(data = rock.scree, aes(x = clusters, y = vaf)) + geom_line() + geom_point()
```

Based on the data set abov, five clusters will be used.

__Cluster Details__

```{r}
rock.train.cluster[[5]]$centers
```

```{r}
rock.centers.data <- as.data.frame(rock.train.cluster[[5]]$centers)
rock.centers.data$group <- 1:5
rock.centers.data$set <- 'train'

rock.centers.data <- rock.centers.data %>% 
    gather(key = 'variable',
           value = 'value',
           factor_key = TRUE,
           -group,
           -set)

ggplot(rock.centers.data, aes(x = variable, y = value, group = group)) + geom_col(position = 'dodge') + 
    facet_grid(~group) + coord_flip() + theme(legend.position = 'bottom')
```

__Test Comparison__

```{r}
rock.test.cluster <- vector("list", length = 15)
rock.test <- rock[-rock.split.index, ]
rock.test$tempo <- (rock.test$tempo-min(rock.test$tempo))/(max(rock.test$tempo)-min(rock.test$tempo))

set.seed(1027)
for (i in 1:15) {
    rock.test.cluster[[i]] <- kmeans(rock.test[, 14:21], centers = as.matrix(rock.train.cluster[[i]]$centers), nstart = 1)
}


for (i in 1:15) {
    rock.scree$holdout.clusters[i] <- i 
    rock.scree$holdout.within.ss[i] <- rock.test.cluster[[i]]$tot.withinss
    rock.scree$holdout.vaf[i] <-  rock.test.cluster[[i]]$betweenss / rock.test.cluster[[i]]$totss
}
```

```{r}
ggplot(data = rock.scree, aes(x = clusters)) + geom_line(aes(y = holdout.vaf), col = 'red') + 
    geom_point(aes(y = holdout.vaf), col = 'red') + geom_line(aes(y = vaf), col = 'black') + 
    geom_point(aes(y = vaf), col = 'black')
```

```{r}
rock.centers.test <- as.data.frame(rock.test.cluster[[5]]$centers)
rock.centers.test$group <- 1:5
rock.centers.test$set <- 'test'

rock.centers.test <- rock.centers.test %>% 
    gather(key = 'variable',
           value = 'value',
           factor_key = TRUE,
           -group,
           -set)

rock.centers.combined <- rbind(rock.centers.data, rock.centers.test)

ggplot(rock.centers.combined, aes(x = variable, y = value, group = interaction(group,set), fill = set)) + 
    geom_col(position = 'dodge') + facet_grid(~group) + coord_flip() + 
    theme(legend.position = 'bottom')
```

Export cluster

```{r}
saveRDS(rock.train.cluster, file = 'rock_clusters.rds')
```

## Electronic Cluster

```{r}
electronic <- music.data[music.data$genre_top == 'Electronic',]

set.seed(1027)
electronic.split.index <- createDataPartition(electronic$track_id, p = 0.632, 
                                  list = F, 
                                  times = 1)

electronic.train.cluster <- vector("list", length = 9)
electronic.train <- electronic[electronic.split.index, ]
electronic.train$tempo <- (electronic.train$tempo-min(electronic.train$tempo))/(max(electronic.train$tempo)-min(electronic.train$tempo))


set.seed(1027)
for (i in 1:15) {
    electronic.train.cluster[[i]] <- kmeans(electronic.train[, 14:21], centers = i, nstart = 100, iter.max = 15)
}


electronic.scree <- data.frame(clusters = rep(NA, 15), within.ss = rep(NA, 15), vaf = rep(NA, 15))
for (i in 1:15) {
    electronic.scree$clusters[i] <- i
    electronic.scree$within.ss[i] <- electronic.train.cluster[[i]]$tot.withinss
    electronic.scree$vaf[i] <-  electronic.train.cluster[[i]]$betweenss /electronic.train.cluster[[i]]$totss
}

ggplot(data = electronic.scree, aes(x = clusters, y = vaf)) + geom_line() + geom_point()
```

Based on the data set abov, five clusters will be used.

__Cluster Details__

```{r}
electronic.train.cluster[[5]]$centers
```

```{r}
electronic.centers.data <- as.data.frame(electronic.train.cluster[[5]]$centers)
electronic.centers.data$group <- 1:5
electronic.centers.data$set <- 'train'

electronic.centers.data <- electronic.centers.data %>% 
    gather(key = 'variable',
           value = 'value',
           factor_key = TRUE,
           -group,
           -set)

ggplot(electronic.centers.data, aes(x = variable, y = value, group = group)) + geom_col(position = 'dodge') + 
    facet_grid(~group) + coord_flip() + theme(legend.position = 'bottom')
```

```{r}
electronic.test.cluster <- vector("list", length = 15)
electronic.test <- electronic[-electronic.split.index, ]
electronic.test$tempo <- (electronic.test$tempo-min(electronic.test$tempo))/(max(electronic.test$tempo)-min(electronic.test$tempo))

set.seed(1027)
for (i in 1:15) {
    electronic.test.cluster[[i]] <- kmeans(electronic.test[, 14:21], centers = as.matrix(electronic.train.cluster[[i]]$centers), nstart = 1)
}


for (i in 1:15) {
    electronic.scree$holdout.clusters[i] <- i 
    electronic.scree$holdout.within.ss[i] <- electronic.test.cluster[[i]]$tot.withinss
    electronic.scree$holdout.vaf[i] <-  electronic.test.cluster[[i]]$betweenss / electronic.test.cluster[[i]]$totss
}
```

```{r}
ggplot(data = electronic.scree, aes(x = clusters)) + geom_line(aes(y = holdout.vaf), col = 'red') + 
    geom_point(aes(y = holdout.vaf), col = 'red') + geom_line(aes(y = vaf), col = 'black') + 
    geom_point(aes(y = vaf), col = 'black')

```

```{r}
electronic.centers.test <- as.data.frame(electronic.test.cluster[[5]]$centers)
electronic.centers.test$group <- 1:5
electronic.centers.test$set <- 'test'

electronic.centers.test <- electronic.centers.test %>% 
    gather(key = 'variable',
           value = 'value',
           factor_key = TRUE,
           -group,
           -set)

electronic.centers.combined <- rbind(electronic.centers.data, electronic.centers.test)

ggplot(electronic.centers.combined, aes(x = variable, y = value, group = interaction(group,set), fill = set)) + 
    geom_col(position = 'dodge') + facet_grid(~group) + coord_flip() + 
    theme(legend.position = 'bottom')
```

Export cluster

```{r}
saveRDS(electronic.train.cluster, file = 'electronic_clusters.rds')
```

## Hip Hop Cluster

```{r}
hiphop <- music.data[music.data$genre_top == 'Hip-Hop',]

set.seed(1027)
hiphop.split.index <- createDataPartition(hiphop$track_id, p = 0.632, 
                                  list = F, 
                                  times = 1)

hiphop.train.cluster <- vector("list", length = 9)
hiphop.train <- hiphop[hiphop.split.index, ]
hiphop.train$tempo <- (hiphop.train$tempo-min(hiphop.train$tempo))/(max(hiphop.train$tempo)-min(hiphop.train$tempo))


set.seed(1027)
for (i in 1:15) {
    hiphop.train.cluster[[i]] <- kmeans(hiphop.train[, 14:21], centers = i, nstart = 100, iter.max = 15)
}


hiphop.scree <- data.frame(clusters = rep(NA, 15), within.ss = rep(NA, 15), vaf = rep(NA, 15))
for (i in 1:15) {
    hiphop.scree$clusters[i] <- i
    hiphop.scree$within.ss[i] <- hiphop.train.cluster[[i]]$tot.withinss
    hiphop.scree$vaf[i] <-  hiphop.train.cluster[[i]]$betweenss / hiphop.train.cluster[[i]]$totss
}

ggplot(data = hiphop.scree, aes(x = clusters, y = vaf)) + geom_line() + geom_point()
```

Based on the data set abov, five clusters will be used.

__Cluster Details__

```{r}
hiphop.train.cluster[[5]]$centers
```

```{r}
hiphop.centers.data <- as.data.frame(hiphop.train.cluster[[5]]$centers)
hiphop.centers.data$group <- 1:5
hiphop.centers.data$set <- 'train'

hiphop.centers.data <- hiphop.centers.data %>% 
    gather(key = 'variable',
           value = 'value',
           factor_key = TRUE,
           -group,
           -set)

ggplot(hiphop.centers.data, aes(x = variable, y = value, group = group)) + geom_col(position = 'dodge') + 
    facet_grid(~group) + coord_flip() + theme(legend.position = 'bottom')
```

```{r}
hiphop.test.cluster <- vector("list", length = 15)
hiphop.test <- hiphop[-hiphop.split.index, ]
hiphop.test$tempo <- (hiphop.test$tempo-min(hiphop.test$tempo))/(max(hiphop.test$tempo)-min(hiphop.test$tempo))

set.seed(1027)
for (i in 1:15) {
    hiphop.test.cluster[[i]] <- kmeans(hiphop.test[, 14:21], centers = as.matrix(hiphop.train.cluster[[i]]$centers), nstart = 1)
}


for (i in 1:15) {
    hiphop.scree$holdout.clusters[i] <- i 
    hiphop.scree$holdout.within.ss[i] <- hiphop.test.cluster[[i]]$tot.withinss
    hiphop.scree$holdout.vaf[i] <-  hiphop.test.cluster[[i]]$betweenss / hiphop.test.cluster[[i]]$totss
}
```

```{r}
ggplot(data = hiphop.scree, aes(x = clusters)) + geom_line(aes(y = holdout.vaf), col = 'red') + 
    geom_point(aes(y = holdout.vaf), col = 'red') + geom_line(aes(y = vaf), col = 'black') + 
    geom_point(aes(y = vaf), col = 'black')

```

```{r}
hiphop.centers.test <- as.data.frame(hiphop.test.cluster[[5]]$centers)
hiphop.centers.test$group <- 1:5
hiphop.centers.test$set <- 'test'

hiphop.centers.test <- hiphop.centers.test %>% 
    gather(key = 'variable',
           value = 'value',
           factor_key = TRUE,
           -group,
           -set)

hiphop.centers.combined <- rbind(hiphop.centers.data, hiphop.centers.test)

ggplot(hiphop.centers.combined, aes(x = variable, y = value, group = interaction(group,set), fill = set)) + 
    geom_col(position = 'dodge') + facet_grid(~group) + coord_flip() + 
    theme(legend.position = 'bottom')
```

Interestingly, the acousticness and instrumentalness seem to be the two driving factors here.

Export cluster

```{r}
saveRDS(hiphop.train.cluster, file = 'hiphop_clusters.rds')
```

## Blues / Folk

```{r}
music.data$genre_top <- recode(music.data$genre_top, 'Folk' = 'Blues/Folk')
music.data$genre_top <- recode(music.data$genre_top, 'Blues' = 'Blues/Folk')
```

```{r}
folk <- music.data[music.data$genre_top == 'Blues/Folk',]

set.seed(1027)
folk.split.index <- createDataPartition(folk$track_id, p = 0.632, 
                                  list = F, 
                                  times = 1)

folk.train.cluster <- vector("list", length = 9)
folk.train <- folk[folk.split.index, ]
folk.train$tempo <- (folk.train$tempo-min(folk.train$tempo))/(max(folk.train$tempo)-min(folk.train$tempo))


set.seed(1027)
for (i in 1:15) {
    folk.train.cluster[[i]] <- kmeans(folk.train[, 14:21], centers = i, nstart = 100, iter.max = 15)
}


folk.scree <- data.frame(clusters = rep(NA, 15), within.ss = rep(NA, 15), vaf = rep(NA, 15))
for (i in 1:15) {
    folk.scree$clusters[i] <- i
    folk.scree$within.ss[i] <- folk.train.cluster[[i]]$tot.withinss
    folk.scree$vaf[i] <-  folk.train.cluster[[i]]$betweenss / folk.train.cluster[[i]]$totss
}

ggplot(data = folk.scree, aes(x = clusters, y = vaf)) + geom_line() + geom_point()
```

Based on the data set abov, five clusters will be used.

__Cluster Details__

```{r}
folk.train.cluster[[5]]$centers
```

```{r}
folk.centers.data <- as.data.frame(folk.train.cluster[[5]]$centers)
folk.centers.data$group <- 1:5
folk.centers.data$set <- 'train'

folk.centers.data <- folk.centers.data %>% 
    gather(key = 'variable',
           value = 'value',
           factor_key = TRUE,
           -group,
           -set)

ggplot(folk.centers.data, aes(x = variable, y = value, group = group)) + geom_col(position = 'dodge') + 
    facet_grid(~group) + coord_flip() + theme(legend.position = 'bottom')
```

```{r}
folk.test.cluster <- vector("list", length = 15)
folk.test <- folk[-folk.split.index, ]
folk.test$tempo <- (folk.test$tempo-min(folk.test$tempo))/(max(folk.test$tempo)-min(folk.test$tempo))

set.seed(1027)
for (i in 1:15) {
    folk.test.cluster[[i]] <- kmeans(folk.test[, 14:21], centers = as.matrix(folk.train.cluster[[i]]$centers), nstart = 1)
}


for (i in 1:15) {
    folk.scree$holdout.clusters[i] <- i 
    folk.scree$holdout.within.ss[i] <- folk.test.cluster[[i]]$tot.withinss
    folk.scree$holdout.vaf[i] <-  folk.test.cluster[[i]]$betweenss / folk.test.cluster[[i]]$totss
}
```

```{r}
ggplot(data = folk.scree, aes(x = clusters)) + geom_line(aes(y = holdout.vaf), col = 'red') + 
    geom_point(aes(y = holdout.vaf), col = 'red') + geom_line(aes(y = vaf), col = 'black') + 
    geom_point(aes(y = vaf), col = 'black')

```

```{r}
folk.centers.test <- as.data.frame(folk.test.cluster[[5]]$centers)
folk.centers.test$group <- 1:5
folk.centers.test$set <- 'test'

folk.centers.test <- folk.centers.test %>% 
    gather(key = 'variable',
           value = 'value',
           factor_key = TRUE,
           -group,
           -set)

folk.centers.combined <- rbind(folk.centers.data, folk.centers.test)

ggplot(folk.centers.combined, aes(x = variable, y = value, group = interaction(group,set), fill = set)) + 
    geom_col(position = 'dodge') + facet_grid(~group) + coord_flip() + 
    theme(legend.position = 'bottom')
```

Export cluster

```{r}
saveRDS(folk.train.cluster, file = 'folk_clusters.rds')
```


## LPA

Latent profile analysis is similar to Latent Class Analysis, but works for quantitative data. We will be testing this using the MClust package.

```{r}
explore_model_fit <- function(df, n_profiles_range = 1:9, model_names = c("EII", "VVI", "EEE", "VVV")) {
    x <- mclustBIC(df, G = n_profiles_range, modelNames = model_names)
    y <- x %>%
        as.data.frame.matrix() %>%
        rownames_to_column("n_profiles") %>%
        rename(`Constrained variance, fixed covariance` = EII, 
               `Freed variance, fixed covariance` = VVI,
               `Constrained variance, constrained covariance` = EEE,
               `Freed variance, freed covariance` = VVV)
    y
}

fit_output <- explore_model_fit(train[, 14:21], n_profiles_range = 1:9)

to_plot <- fit_output %>%
    gather(`Covariance matrix structure`, val, -n_profiles) %>% 
    mutate(`Covariance matrix structure` = as.factor(`Covariance matrix structure`),
           val = abs(val)) # this is to make the BIC values positive (to align with more common formula / interpretation of BIC)

to_plot$`Covariance matrix structure` <- fct_relevel(to_plot$`Covariance matrix structure`,
                                                     "Constrained variance, fixed covariance",
                                                     "Freed variance, fixed covariance",
                                                     "Constrained variance, constrained covariance",
                                                     "Freed variance, freed covariance")


ggplot(to_plot, aes(x = n_profiles, y = val, color = `Covariance matrix structure`, group = `Covariance matrix structure`)) +
    geom_line() +
    geom_point() +
    ylab("BIC")

Music.LPA <- Mclust(train[, 14:21], G = 6, modelNames = 'EII')
summary(Music.LPA)
lpa.center <- t(Music.LPA$parameters$mean)
row.order <- c( 3, 4, 6, 2, 1, 5)
lpa.center <- as.data.frame(lpa.center[row.order, , drop = F])
```

```{r}
lpa.center$group <- 1:6
lpa.center$set <- 'lpa'

lpa.center <- lpa.center %>% 
    gather(key = 'variable',
           value = 'value',
           factor_key = TRUE,
           -group,
           -set)

centers.clust.lpa <- rbind(centers.data, lpa.center)

ggplot(centers.clust.lpa, aes(x = variable, y = value, group = interaction(group,set), fill = set)) + 
    geom_col(position = 'dodge') + geom_col(position = 'dodge') + facet_grid(~group) + coord_flip() + 
    theme(legend.position = 'bottom')
```

```{r}
table(Music.LPA$classification)
table(music.k.means[[6]]$cluster)

table(Music.LPA$classification, music.k.means[[6]]$cluster)
```
