---
title: "Data Mining Project: Genre Prediction"
author: "Stephen Gianfortoni, Anna Whelan, and Scott Shepard"
date: "August 2, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r, include=F}
library(caret)
library(tidyverse)
```

## Load Data

International and Experimental songs were dropped. They both had low volume, and are defined by features besides their inherent musical features, so they provided instrinsic issues to classification. An alternative sample set that had an even number of each genre would be prefereable, but unfortunately we did not have this data.

```{r}
set.seed(1)
datapath <- 'C:/Users/gianf/Documents/MS Analytics/Data Mining/'
full.data <- read.csv(paste0(datapath, 'Project_Data.csv'))

empty.genre <- which(full.data$genre_top=="")
data.empty.genre <- full.data[full.data$genre_top == '', ]

international.genre <- which(full.data$genre_top=="International")
experimental.genre <- which(full.data$genre_top=="Experimental")
blues.genre <- which(full.data$genre_top=="Blues")
full.data$genre_top[blues.genre] <- "Folk"
full.data <- full.data[-c(empty.genre, international.genre, experimental.genre),]

full.data <- droplevels(full.data)
levels(full.data$genre_top)[3] <- "Folk/Blues"

samp <- sample(1:nrow(full.data), size=0.7*nrow(full.data), replace=FALSE)

d.train <- full.data[samp,]
d.test <- full.data[-samp,]

d.train.trim <- d.train[, c(11, 14, 15, 16, 17, 18, 19, 20, 21)]
d.test.trim <- d.test[, c(11, 14, 15, 16, 17, 18, 19, 20, 21)]

```


## Decision Tree

```{r}
library(rpart)
#d.train.trim.2 <- d.train.trim[, -c(6, 8, 9)]

train.tree <- rpart(d.train.trim, control=rpart.control(cp=0, minsplit=25, xval=20, maxsurrogate=0))

#plot(train.tree, main="Classification Tree: Music Data", col=3, compress=TRUE, branch=0.2, uniform=TRUE)
#text(train.tree, cex=0.8, col=4, fancy=TRUE, fwidth=0.1, fheight=0.1, bg=c(5))

#plotcp(train.tree, minline=TRUE)
printcp(train.tree)

```

```{r}
#pruned tree
train.tree.pruned <- rpart(d.train.trim, control=rpart.control(cp=9.6360e-04, minsplit=25, xval=20, maxsurrogate=0))

#plot(train.tree.pruned, main="Classification Tree: Music Data", col=3, compress=TRUE, branch=0.9, uniform=TRUE)
#text(train.tree.pruned, cex=0.8, col=4, fancy=TRUE, fwidth=0.1, fheight=0.1, bg=c(5))

```

__Train Accuracy__

```{r}

conf.train <- round(prop.table(table(d.train.trim$genre_top, predict(train.tree.pruned, type="class")),1),2)
train.accuracy <- diag(conf.train)

conf.train
train.accuracy
```

__Test Accuracy__

```{r}
test.tree <- predict(train.tree.pruned, newdata = d.test, type = 'class')

conf.test <- round(prop.table(table(d.test.trim$genre_top, test.tree),1),2)
test.accuracy <- diag(conf.test)

conf.test
test.accuracy
```

## Prediction of unlabeled genres

```{r}
#predict genre for empty

data.empty.genre$predicted.genre <- predict(train.tree.pruned, newdata = data.empty.genre, type="class")

summary(data.empty.genre$predicted.genre)

rock.pred <- data.empty.genre[data.empty.genre$predicted.genre == 'Rock',]
rock.pred$tempo <- (rock.pred$tempo-min(rock.pred$tempo))/(max(rock.pred$tempo)-min(rock.pred$tempo))

hiphop.pred <- data.empty.genre[data.empty.genre$predicted.genre == 'Hip-Hop',]
hiphop.pred$tempo <- (hiphop.pred$tempo-min(hiphop.pred$tempo))/(max(hiphop.pred$tempo)-min(hiphop.pred$tempo))

electronic.pred <- data.empty.genre[data.empty.genre$predicted.genre == 'Electronic',]
electronic.pred$tempo <- (electronic.pred$tempo-min(electronic.pred$tempo))/(max(electronic.pred$tempo)-min(electronic.pred$tempo))

folk.pred <- data.empty.genre[data.empty.genre$predicted.genre == 'Folk/Blues',]
folk.pred$tempo <- (folk.pred$tempo-min(folk.pred$tempo))/(max(folk.pred$tempo)-min(folk.pred$tempo))


```

Importing the clusters allows us to see if the clusters match.

Rock:

```{r}
rock.clusters <- readRDS(file = 'rock_clusters.rds')
```

```{r}
rock.pred.cluster <- kmeans(rock.pred[, 14:21], centers = as.matrix(rock.clusters[[5]]$centers), nstart = 1)

rock.clusters[[5]]$centers

rock.pred.cluster$centers
```

Plot center comparisons

```{r}

rock.train <- data.frame(rock.clusters[[5]]$centers)
rock.train$group <- 1:5
rock.train$set <- 'train'
rock.train <- rock.train %>% 
    gather(key = 'variable',
           value = 'value',
           factor_key = TRUE,
           -group,
           -set)

rock.unk <- data.frame(rock.pred.cluster$centers)
rock.unk$group <- 1:5
rock.unk$set <- 'unknown'
rock.unk <- rock.unk %>% 
    gather(key = 'variable',
           value = 'value',
           factor_key = TRUE,
           -group,
           -set)

rock.graph <- rbind(rock.unk, rock.train)

ggplot(rock.graph, aes(x = variable, y = value, group = interaction(group,set), fill = set)) + 
    geom_col(position = 'dodge') + geom_col(position = 'dodge') + facet_grid(~group) + coord_flip() + 
    theme(legend.position = 'bottom')

```

Hip Hop

```{r}
hiphop.clusters <- readRDS(file = 'hiphop_clusters.rds')
```

```{r}
hiphop.pred.cluster <- kmeans(hiphop.pred[, 14:21], centers = as.matrix(hiphop.clusters[[5]]$centers), nstart = 1)

hiphop.clusters[[5]]$centers

hiphop.pred.cluster$centers
```

Plot center comparisons

```{r}

hiphop.train <- data.frame(hiphop.clusters[[5]]$centers)
hiphop.train$group <- 1:5
hiphop.train$set <- 'train'
hiphop.train <- hiphop.train %>% 
    gather(key = 'variable',
           value = 'value',
           factor_key = TRUE,
           -group,
           -set)

hiphop.unk <- data.frame(hiphop.pred.cluster$centers)
hiphop.unk$group <- 1:5
hiphop.unk$set <- 'unknown'
hiphop.unk <- hiphop.unk %>% 
    gather(key = 'variable',
           value = 'value',
           factor_key = TRUE,
           -group,
           -set)

hiphop.graph <- rbind(hiphop.unk, hiphop.train)

ggplot(hiphop.graph, aes(x = variable, y = value, group = interaction(group,set), fill = set)) + 
    geom_col(position = 'dodge') + geom_col(position = 'dodge') + facet_grid(~group) + coord_flip() + 
    theme(legend.position = 'bottom')

```

Notice that there are significant differences in clusters 1 and 5. These may be classified incorrectly.

Electronic

```{r}
electronic.clusters <- readRDS(file = 'electronic_clusters.rds')
```

```{r}
electronic.pred.cluster <- kmeans(electronic.pred[, 14:21], centers = as.matrix(electronic.clusters[[5]]$centers), nstart = 1)

electronic.clusters[[5]]$centers

electronic.pred.cluster$centers
```

Plot center comparisons

```{r}

electronic.train <- data.frame(electronic.clusters[[5]]$centers)
electronic.train$group <- 1:5
electronic.train$set <- 'train'
electronic.train <- electronic.train %>% 
    gather(key = 'variable',
           value = 'value',
           factor_key = TRUE,
           -group,
           -set)

electronic.unk <- data.frame(electronic.pred.cluster$centers)
electronic.unk$group <- 1:5
electronic.unk$set <- 'unknown'
electronic.unk <- electronic.unk %>% 
    gather(key = 'variable',
           value = 'value',
           factor_key = TRUE,
           -group,
           -set)

electronic.graph <- rbind(electronic.unk, electronic.train)

ggplot(electronic.graph, aes(x = variable, y = value, group = interaction(group,set), fill = set)) + 
    geom_col(position = 'dodge') + geom_col(position = 'dodge') + facet_grid(~group) + coord_flip() + 
    theme(legend.position = 'bottom')

```

Folk

```{r}
folk.clusters <- readRDS(file = 'folk_clusters.rds')
```

The clusters break down below? Maybe influx of other high accoustice genres? Or the classifier only identifies Folk as being highly acoustic.

```{r}
set.seed(1027)
folk.pred.cluster <- kmeans(folk.pred[, 14:21], centers = 5, nstart = 10)

folk.clusters[[5]]$centers

folk.pred.cluster$centers
```

Plot center comparisons

```{r}

folk.train <- data.frame(folk.clusters[[5]]$centers)
folk.train$group <- 1:5
folk.train$set <- 'train'
folk.train <- folk.train %>% 
    gather(key = 'variable',
           value = 'value',
           factor_key = TRUE,
           -group,
           -set)

folk.unk <- data.frame(folk.pred.cluster$centers)
folk.unk$group <- 1:5
folk.unk$set <- 'unknown'
folk.unk <- folk.unk %>% 
    gather(key = 'variable',
           value = 'value',
           factor_key = TRUE,
           -group,
           -set)

folk.graph <- rbind(folk.unk, folk.train)

ggplot(folk.graph, aes(x = variable, y = value, group = interaction(group,set), fill = set)) + 
    geom_col(position = 'dodge') + geom_col(position = 'dodge') + facet_grid(~group) + coord_flip() + 
    theme(legend.position = 'bottom')

```
